---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Fit SSFs

Here we fit SSFs to the data.

## Load libraries and prepare files

```{r}
# libs for data
library(dplyr)
library(tidyr)
library(purrr)

# libs for messages
library(glue)

# library for SSF
library(amt)
```

```{r}
# load data saved as Rds object
load("data/processed/data_for_ssf.Rds")
```

## Fit candidate SSF

Here, we fit a candidate SSF for step length, NDVI, landcover, the interaction of step length and NDVI and the interaction of step length and landcover. All environmental covariates are taken at the end point of the steps, i.e., to determine which conditions the birds prefer to move to (see Signer et al. 2019).

```{r}
# make a copy of the data, renaming data to ssf
ssfs = rename(tracks, ssf = "data")
```

```{r}
# specify the model formula
ssf_formula = "case_ ~ ndvi + landcover + strata(step_id_)"

# add formula to dataframe
ssfs = mutate(
  ssfs,
  ssf_formula = ssf_formula
)
```


```{r}
# sink output to log
sink(file = "data/log_ssf_fit.log")

ssf_comment = "Trial SSF log writing"

# first print date
print(
  glue("
    SSF fit log from {Sys.time()}
    
    comment: {ssf_comment}
    
  "
  )
)

# fit ssfs overwriting the track data
ssfs = mutate(
  ssfs,
  ssf = imap(ssf, function(.x, .y) {
    
    # print some information about the data -- which row, which id, treatment
    print(
      glue("sp = {ssfs$sp[.y]}; \\
            tag = {ssfs$TAG_ID[.y]}; treatment = {ssfs$treat[.y]}
            track date = {ssfs$date[.y]}
            row in dataframe = {.y}
            formula = {ssfs$ssf_formula[.y]}
            start time: {Sys.time()}
           ")
    )
    
    # pass the model fitting through a try catch
    # this allows errors to be logged and prevents the process from
    # breaking when there is an error
    tryCatch(
      
      # the model fitting expression to evaluate
      expr = {
        
        # this is the SSF fit function from amt
        ssf_fit = 
          .x %>% 
          amt::fit_issf(
            formula = as.formula(ssfs$ssf_formula[.y])
          )
        
        # print success message to the processing log
        print(
          glue("
            SUCCESS at time: {Sys.time()}
            ---
            
            ")
        )
        
        # return the model fit object
        ssf_fit
      }, 
      error = function(e) {
        
        # print an error message with information about the data
        print(
          glue("FAILED at time: {Sys.time()}")
        )
        # print the error message itself
        print(e)
        
        # a trailing blank line
        print(
          glue("
            ---
            
            ")
        )
        
        # return null if there is an error
        NULL
      }
    )
  })
)

# end sinking to log
sink()
```

### Save SSF data

```{r}
save(ssfs, file = "data/results/data_fitted_ssfs.Rds")
```

